---
title: "Data cleaning & transformation with dplyr"
output: github_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, warning = FALSE, message = FALSE)
```

## GitHub Documents

This is an R Markdown format used for publishing markdown documents to GitHub. When you click the **Knit** button all R code chunks are run and a markdown file (.md) suitable for publishing to GitHub is generated.

## R directories set ups

```{r}
dir.create("data")
dir.create("data_output")
dir.create("m_output")
```

## Load the libraries

```{r}
library(ggplot2)
library(tidyverse)
library(dplyr)
library(tidyr)
library(lubridate)
```

## Upload dataset

```{r}
trains <- read_csv("data/trains.csv")
trains
write.csv (trains, "trains.csv")
```
## Renaming variables

```{r}
names(trains)
trains2 <- rename(trains,
                index = Statistics,
                rdate = REF_DATE)

trains2
```
## Recoding values

```{r}
unique(trains2$Companies)
unique(trains2$index)
unique(trains2$rdate)
unique(trains2$Commodities)
trains2$UOM
trains2$Companies <- recode (trains2$Companies, CSXT = "CSX")
trains2$index <- recode (trains2$index, 
                         "Average train speed" = "avspeed",
                         "Weekly average dwell time at origin" = "w ave", 
                         "Weekly average number of trains holding per day (crew)" = "CREW w.avetr", 
                         "Weekly average number of trains holding per day (locomotive)" = "loc w.avetr",
                         "Weekly average number of trains holding per day (other)" = "other w.avetr",
                         "Weekly average number of trains holding per day (all)" = "all w.avetr",
                         "Weekly average number of loaded cars in revenue service that have not moved 48 hours or greater" = "w.avecarl 48h",
                         "Weekly average number of empty cars in revenue service that have not moved 48 hours or greater" = "w.avecare-48h")
trains2$UOM <- recode(trains2$UOM, "Kilometres per hour" = "km/h", 
                      "Hour" = "h", 
                      "Number" = "n#")
                           
trains2
```

## Subsetting dataframes

```{r}
CN_oil <- filter(trains2, Companies == "Canadian National", Commodities == "Crude oil")
CN_oil
```

## How many items were removed?

```{r}
nrow(trains2) - nrow(CN_oil)
```
# Selecting variables
## specify the variables to keep by name

```{r}
tr_comp <- dplyr::select(trains2, rdate, Companies, Commodities, index, VALUE, UOM_ID, DGUID)
tr_comp
```

## specify the variables you want to remove with a -

```{r}
tr_comp_r <- select(trains2, -UOM_ID, -DGUID)
tr_comp_r
```
## reorder columns, combined with everything()

```{r}
trains2_reordered <- select(trains2, Companies, rdate, everything())
trains2_reordered
```

# Order data
## The arrange() function in the dplyr package allows you to sort your data by alphabetical or numerical order.

```{r}
trains2_arrange <- arrange(trains2, Companies)
trains2_arrange
```
## use desc() to sort a variable in descending order

```{r}
trains2_compbydatedesc <- arrange(trains2, desc(rdate))
trains2_compbydatedesc
```

## order data based on multiple variables (e.g. sort first by date, then by value)

```{r}
trainsadateValue <- arrange(trains2, desc(rdate), desc(VALUE))
```
# Creating new variables
## the mutate() function allows to create new variables. Here, we use the str_sub() function from the stringr package to extract the first character of the callnumber variable (the call number class) and put it into a new column called call_class.

```{r}
trains2_datesplit <- trains2 %>%
  dplyr::mutate(trains,
                year = lubridate::year(rdate), 
                month = lubridate::month(rdate), 
                day = lubridate::day(rdate)) %>% 
          select(-rdate, year, month, day, Companies, Commodities, index, UOM, VALUE)
```

# Putting it all together with %>%
# the Pipe Operator %>% is loaded with the tidyverse. It takes the output of one statement and makes it the input of the next statement.
# it is used with Ctrl + Shift + M if you have a PC or Cmd + Shift + M if you have a Mac.

```{r}
trains_compl_search <- trains2_datesplit %>% 
  filter(Companies == "Canadian National", Commodities == "Crude oil", year == "2019") %>% 
  select(month, index, VALUE, -year) %>% 
  drop_na() %>% 
  group_by(month) %>% 
  summarize(mean_index = mean(VALUE))
trains_compl_search
```
## Create a vector base for the graphs

```{r}
cleaned_t2<-trains2 %>%
  select(rdate, index, Commodities, Companies, VALUE ) %>%
  drop_na()%>%
  arrange(rdate)
cleaned_t2
```

# scatterplot

```{r}
ggplotbase <-cleaned_t2 %>%
  filter(Companies == "Canadian Pacific", index == "all w.avetr") %>%
  ggplot(mapping = aes(x = rdate, y = VALUE, color = Commodities)) 
ggplotbase +  geom_point(size = 2) +
  labs(title = "Weekly average number of trains holding per day",
       subtitle = "Canadian Pacific",
       y = "VALUE",
       x = "rdate") + theme_bw(base_size = 15)
```
# bar chart

```{r}
ggplotbase <- cleaned_t2 %>%
  filter(Companies == "Canadian Pacific", index == "all w.avetr") %>%
  ggplot()
ggplotbase +geom_bar(aes(x = rdate, y = VALUE, fill = Commodities), position = position_stack(reverse = TRUE), stat="identity") +
  labs(title = "Weekly average number of trains holding per day: Canadian Pacific", 
       y = "VALUE",
       x = "rdate") + 
  theme(legend.position = "top", plot.title = element_text(hjust = 0.5))+
  scale_fill_discrete(breaks=c("System-wide","Other commodities","Manifest", "Intermodal", "Grain", "Ethanol", "Crude oil", "Coal"< "Automotive"))
```

## Write an output

```{r}
write.csv(cleaned_t2, "trains_output_final.csv", row.names = F)
```